---
title: "Preprocess metadata"
format:
  html:
    embed-resources: true
---

```{r}
#| message: false
library(haven)
library(readxl)
library(tidyverse)
```

## Check metadata information files

We have been sent two metadata files that are stored in `data`:

```{r}
id_group_pbmc <- read_xlsx("../data/id_group_PBMC.xlsx")
PRETTY_metadata <- read_dta("../data/PRETTY_metadata.dta")
```

-   `id_group_PBMC.xlsx`: Database with a list with 164 aliquots of the PBMC samples that were requested to the biobank. It contains 6 columns: `id_sample`, `group`, `alicuota_pbmc`, `IDTUBO`, `millioncells`, `Caja`, and `Fila`. This was used to perform the RNA-seq experiment.

-   `PRETTY_metadata.dta`: Final **correct and complete** database. It is the same as `DEFINITIVOS CON MUESTRA BIOBANCO.dta`, with 187 samples and 68 variables.

We will create a final file named `PRETTY_metadata.tsv`, which will be used throughout the project.

## Final metadata database

We will use the PRETTY_metadata.dta database and complete it with the ids for the plasma proteomics and RNA-seq samples, so that we have all information together in a single file.

We change the codification of the group variable so that `1=Event`, and `0=Control` and we also change the `NA` values in the `death_all` variables to be `0=No`.

```{r}
PRETTY_metadata <- PRETTY_metadata |> 
  mutate(
    group = if_else(group == 0, 1, 0),
    death_all = if_else(is.na(death_all), 0, death_all)
  )
```

We add the Olink proteomics IDs:

```{r}
PRETTY_metadata$olink_id <- PRETTY_metadata$id_muestra
```

We add the RNA-seq IDs:

```{r}
id_group_pbmc <- id_group_pbmc |> mutate(rnaseq_id = gsub("-", "_", IDTUBO)) |> 
  select(id_sample, rnaseq_id)
PRETTY_metadata <- PRETTY_metadata |>
  left_join(id_group_pbmc, by = c("id_muestra" = "id_sample"))
```

Now, we are adding a specific number for each sample, so we can keep track of the samples more easily.

```{r}
PRETTY_metadata <- PRETTY_metadata |>
  mutate(sid = paste0("PTY_", sprintf("%03d", row_number())))
```

Finally, we order it so that we have the main variables of interest at the beginning of the table:

```{r}
PRETTY_metadata <- PRETTY_metadata |> select(sid, id, olink_id, rnaseq_id, group, nadm, mace, death_all, everything())
```

## Save to file

```{r}
# PRETTY_metadata |> write_tsv("../data/PRETTY_metadata.tsv")
```
