---
title: "Preprocess and QC Olink"
format:
  html:
    df-print: paged
    embed-resources: true
---

```{r}
#| message: false
library(OlinkAnalyze)
library(readxl)
library(tidyverse)
```

In the Olink report, we have the following QC relevant information, that we will take as starting point:

-   Olink Panel: Explore 384 Inflammation

-   Normalization method: Intensity normalization (except PNLIPRP2)

-   Number of samples: 187

-   Samples passed QC: 171

-   Number of proteins: 368

-   Number of detected proteins: 338

## Read NPX data

```{r}
pdata <- read_NPX("../data/Q-03707_Serrano_NPX_2022-09-29.csv")
```

Add information about the groups:

```{r}
#| message: false
mdata_groups <- read_tsv("../data/PRETTY_metadata.tsv") |> 
  select(sid, olink_id, group, nadm, mace, death_all) |> 
  mutate(olink_id = as.character(olink_id))
pdata <- pdata |> left_join(mdata_groups, by = c("SampleID" = "olink_id"))
```

## Number of measurements

First, we count the number of measurements in each group:

```{r}
pdata |> count(group) |> mutate(pts = n / 368)
```

We have 95 patients in the control group and 92 patients in the events group. There are 3 patients that do not have a correct PS-matched pair in the control group. We will remove this samples from further analyses.

```{r}
#| message: false
mdata_ids <- read_tsv("../data/PRETTY_metadata.tsv") |> select(sid, id)
```

```{r}
id_matching <- read_xlsx("../data/id_matching_controls_cases.xlsx")
id_matching_pairs <- id_matching |> 
  bind_rows(rename(id_matching, ID_event = ID_control, ID_control = ID_event)) |> 
  filter(!is.na(ID_control)) |>
  rename(id = ID_event, id_pair = ID_control) |> 
  left_join(mdata_ids, by = c("id_pair" = "id")) |> 
  select(sid, id, id_pair)
```

## Samples quality control

### Filter 1: QC Warning

Remove control samples:

```{r}
pdata <- pdata |> filter(!grepl("CONTROL", SampleID))
```

Remove samples with Warning QC status:

```{r}
sample_warn <- pdata |> 
  filter(QC_Warning == "WARN") |> 
  count(sid) |> 
  mutate(percentage = round(n/n_distinct(pdata$Assay)*100)) |> 
  arrange(desc(percentage))
sample_warn_ids <- sample_warn |> pull(sid)
sample_warn
```

Check pairs of these samples:

```{r}
# sample_warn_remove_ids <- mdata_groups |> filter(sid %in% sample_warn_ids) |> pull(sid)
sample_warn_ids_pairs <- id_matching_pairs |> filter(sid %in% sample_warn_ids) |> pull(id)
sample_warn_ids_pairs_sid <- id_matching_pairs |> filter(id_pair %in% sample_warn_ids_pairs) |> pull(sid)
sample_warn_remove_all <- unique(c(sample_warn_ids, sample_warn_ids_pairs_sid))
```

```{r}
# Remove samples
pdata_filter1 <- pdata |> filter(!sid %in% sample_warn_remove_all)
```

This is the total number of remaining samples:

```{r}
pdata_filter1 |> count(group) |> mutate(pts = n / 368)
```

### Filter 2: Outliers

Identify outliers:

```{r}
qc <- olink_qc_plot(pdata_filter1)
outliers <- qc$data |> filter(Outlier == 1) |> pull(SampleID)
outliers_sid <- pdata |> select(SampleID, sid) |> distinct() |>  filter(SampleID %in% outliers) |> pull(sid)
```

5 outliers were identified. We are going to remove these 5 and their paired participants.

Identify pairs of these samples:

```{r}
outliers_ids_pairs <- id_matching_pairs |> filter(sid %in% outliers_sid) |> pull(id)
outliers_ids_pairs_sid <- id_matching_pairs |> filter(id_pair %in% outliers_ids_pairs) |> pull(sid)
outliers_remove_all <- unique(c(outliers_sid, outliers_ids_pairs_sid))
```

Remove:

```{r}
pdata_filter2 <- pdata_filter1 |> filter(!sid %in% outliers_remove_all)
```

This is the total number of remaining samples:

```{r}
pdata_filter2 |> count(group) |> mutate(pts = n / 368)
```

### Filter 3: Unmatched samples

Identify unmatched samples:

```{r}
unmatched_sid <- id_matching_pairs |> filter(is.na(id)) |> pull(sid)
```

Remove:

```{r}
pdata_filter3 <- pdata_filter2 |> filter(!sid %in% unmatched_sid)
```

This is the total number of remaining samples:

```{r}
pdata_filter3 |> count(group) |> mutate(pts = n / 368)
```

## Assays quality control

### Filter 4: Missingness

Identify assays with high missingness:

```{r}
assays_todel <- pdata_filter3 |> 
  mutate(Missing = ifelse(NPX > LOD, FALSE, TRUE)) |> 
  group_by(Assay, group) |> 
  summarise(Missing = round(mean(Missing), 4 ) > 0.25) |> 
  group_by(Assay) |>
  summarise(Missing_status = all(Missing)) |> 
  filter(Missing_status == TRUE) |>
  pull(Assay)
```

Remove:

```{r}
pdata_filter4 <- pdata_filter3 |> filter(!Assay %in% assays_todel)
```

### Filter 5: Assay Warning

Count assay warnings:

```{r}
pdata_filter4 |> count(Assay_Warning)
```

There are no flagged assays.

## Final dataset

```{r}
pdata_final <- pdata_filter4
```

The final dataset contains 138 samples and 326 proteins, distributed in the following groups:

```{r}
pdata_final |> select(sid, group) |> distinct() |> count(group)
```

```{r}
pdata_final |> select(sid, nadm) |> distinct() |> count(nadm)
```

```{r}
pdata_final |> select(sid, mace) |> distinct() |> count(mace)
```

```{r}
pdata_final |> select(sid, death_all) |> distinct() |> count(death_all)
```

Save to file:

```{r}
# pdata_final |> write_tsv("../data/PRETTY_NPX_138samp_326prot_QCed.tsv")
```
