---
title: "Survival analysis"
format:
  html:
    df-print: paged
    embed-resources: true
---

```{r}
#| message: false
library(broom)
library(ggpubr)
library(ggsurvfit)
library(patchwork)
library(pROC)
library(survival)
library(survminer)
library(tidyverse)
```

# Kaplan-Meier curves

## Retrieve data

```{r}
#| message: false
pdata <- read_tsv("../data/PRETTY_NPX_138samp_326prot_QCed.tsv") |> 
    mutate(
      group = factor(group, levels = c(1, 0)),
      nadm = factor(nadm, levels = c(1, 0)),
      mace = factor(mace, levels = c(1, 0)),
      death_all = factor(death_all, levels = c(1, 0))
    )
```

## Preparing the data

TO do this, we need a table with:

Group \| Quartile for the protein \| Evento (yes/no) \| T0 \| Tf \| Time to event

Protein data:

```{r}
#| message: false
pdata <- read_tsv("../data/PRETTY_NPX_138samp_326prot_QCed.tsv") |> 
    mutate(
      group = factor(group, levels = c(1, 0)),
      nadm = factor(nadm, levels = c(1, 0)),
      mace = factor(mace, levels = c(1, 0)),
      death_all = factor(death_all, levels = c(1, 0))
    )
```

Metadata:

```{r}
#| message: false
mdata <- read_tsv("../data/PRETTY_metadata.tsv") #|>
  # select(-c("group", "nadm", "mace", "death_all"))
mdata$nadm <- as.factor(mdata$nadm)
mdata$mace <- as.factor(mdata$mace)
mdata$death_all <- as.factor(mdata$death_all)
```

Time data:

```{r}
time_data <- mdata |> 
  dplyr::select(sid, group, art_endfu, fechaENOS, f_muestra24) |>
  mutate(time = ifelse(group == 1,
                       floor(interval(f_muestra24, fechaENOS) / months(1)),
                       floor(interval(f_muestra24, art_endfu) / months(1)))) |> 
  mutate(time_trunc = pmin(time, 60),
         group_trunc = ifelse(time > 60, 0, group))
```

# ELISA cutoffs

## CCL13

```{r}
#| message: false
ccl13_elisa <- read_tsv("../data/Results_CCL13_v2.tsv", locale = locale("es", decimal_mark = ",")) |> 
  dplyr::select("Sample ID", "Mean Concentration [pg/ml]") |> rename_with(~ c("SampleID", "mean_conc")) |>
  left_join((mdata |> dplyr::select(sid, olink_id, group, nadm, mace, death_all)), by = c("SampleID" = "olink_id")) |> 
  mutate(quantile = factor(ntile(mean_conc, 4))) |> 
  left_join(time_data, by = c("sid", "group")) |> 
  filter(time > 0) |> 
  mutate(group = as.factor(group))
```

```{r}
roc_res <- roc(ccl13_elisa$group, ccl13_elisa$mean_conc, direction = "<")  # use direction = "<" if higher values = higher risk
plot(roc_res, print.auc = TRUE)
```

```{r}
opt_cutoff <- coords(roc_res, "best", best.method = "youden", transpose = FALSE)
opt_cutoff
```

```{r}
survfit2(Surv(time, group) ~ quantile, data = ccl13_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
ccl13_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = ccl13_elisa)
summary(ccl13_cox)
```

```{r}
quantile(ccl13_elisa$mean_conc, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
```

```{r}
threshold_time <- 12*5  # in months
ccl13_binary <- ccl13_elisa |> 
  mutate(
    event_within_1y = ifelse(group == 1 & time <= threshold_time, 1, 0),
    test_positive = ifelse(quantile %in% c(3, 4), 1, 0)
  )
```

```{r}
table(ccl13_binary$test_positive, ccl13_binary$event_within_1y)
```

```{r}
TP <- sum(ccl13_binary$test_positive == 1 & ccl13_binary$event_within_1y == 1, na.rm = TRUE)
FP <- sum(ccl13_binary$test_positive == 1 & ccl13_binary$event_within_1y == 0, na.rm = TRUE)
TN <- sum(ccl13_binary$test_positive == 0 & ccl13_binary$event_within_1y == 0, na.rm = TRUE)
FN <- sum(ccl13_binary$test_positive == 0 & ccl13_binary$event_within_1y == 1, na.rm = TRUE)
```

```{r}
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
PPV <- TP / (TP + FP)
NPV <- TN / (TN + FN)

results <- c(Sensitivity = sensitivity,
             Specificity = specificity,
             VPP = PPV,
             VPN = NPV)

round(results, 3)
```

5 years:

```{r}
survfit2(Surv(time_trunc, as.factor(group_trunc)) ~ quantile, data = ccl13_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
summary(coxph(Surv(time_trunc, group_trunc) ~ quantile, data = ccl13_elisa))
```

### Neoplasia

```{r}
survfit2(Surv(time, nadm) ~ quantile, data = ccl13_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### CV event

```{r}
survfit2(Surv(time, mace) ~ quantile, data = ccl13_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Death

```{r}
survfit2(Surv(time, death_all) ~ quantile, data = ccl13_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Hazard ratios, CI, and p-values for Tables

```{r}
ccl13_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = ccl13_elisa)
tidy(ccl13_cox, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
ccl13_cox_nadm <- coxph(Surv(time, as.numeric(nadm)) ~ quantile, data = ccl13_elisa)
tidy(ccl13_cox_nadm, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
ccl13_cox_mace <- coxph(Surv(time, as.numeric(mace)) ~ quantile, data = ccl13_elisa)
tidy(ccl13_cox_mace, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
ccl13_cox_death <- coxph(Surv(time, as.numeric(death_all)) ~ quantile, data = ccl13_elisa)
tidy(ccl13_cox_death, conf.int = TRUE, exponentiate = TRUE)
```

### No. with event for Tables

```{r}
ccl13_elisa |> group_by(quantile) |> dplyr::count(group) |> filter(group == 1)
```

```{r}
ccl13_elisa |> group_by(quantile) |> dplyr::count(nadm) |> filter(nadm == 1)
```

```{r}
ccl13_elisa |> group_by(quantile) |> dplyr::count(mace) |> filter(mace == 1)
```

```{r}
ccl13_elisa |> group_by(quantile) |> dplyr::count(death_all) |> filter(death_all == 1)
```

## DAPP1

```{r}
#| message: false
dapp1_elisa <- read_tsv("../data/Results_DAPP1_v2.tsv", locale = locale("es", decimal_mark = ",")) |> dplyr::select("Sample ID", "Mean Concentration [ng/ml]") |> rename_with(~ c("SampleID", "mean_conc")) |>
  left_join((mdata |> dplyr::select(sid, olink_id, group, nadm, mace, death_all)), by = c("SampleID" = "olink_id")) |> 
  mutate(quantile = factor(ntile(mean_conc, 4))) |> 
  left_join(time_data, by = c("sid", "group")) |> 
  filter(time > 0) |> 
  mutate(group = as.factor(group))
```

```{r}
roc_res <- roc(dapp1_elisa$group, dapp1_elisa$mean_conc, direction = ">")  # use direction = "<" if higher values = higher risk
plot(roc_res, print.auc = TRUE)
```

```{r}
opt_cutoff <- coords(roc_res, "best", best.method = "youden", transpose = FALSE)
opt_cutoff
```

```{r}
survfit2(Surv(time, group) ~ quantile, data = dapp1_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
dapp1_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = dapp1_elisa)
summary(dapp1_cox)
```

```{r}
quantile(dapp1_elisa$mean_conc, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
```

```{r}
threshold_time <- 12*5  # in months
dapp1_binary <- dapp1_elisa |> 
  mutate(
    event_within_1y = ifelse(group == 1 & time <= threshold_time, 1, 0),
    test_positive = ifelse(quantile %in% c(3, 4), 1, 0)
  )
```

```{r}
table(dapp1_binary$test_positive, dapp1_binary$event_within_1y)
```

```{r}
TP <- sum(dapp1_binary$test_positive == 1 & dapp1_binary$event_within_1y == 1, na.rm = TRUE)
FP <- sum(dapp1_binary$test_positive == 1 & dapp1_binary$event_within_1y == 0, na.rm = TRUE)
TN <- sum(dapp1_binary$test_positive == 0 & dapp1_binary$event_within_1y == 0, na.rm = TRUE)
FN <- sum(dapp1_binary$test_positive == 0 & dapp1_binary$event_within_1y == 1, na.rm = TRUE)
```

```{r}
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
PPV <- TP / (TP + FP)
NPV <- TN / (TN + FN)

results <- c(Sensitivity = sensitivity,
             Specificity = specificity,
             VPP = PPV,
             VPN = NPV)

round(results, 3)
```

5 years:

```{r}
survfit2(Surv(time_trunc, as.factor(group_trunc)) ~ quantile, data = dapp1_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
summary(coxph(Surv(time_trunc, group_trunc) ~ quantile, data = dapp1_elisa))
```

### Neoplasia

```{r}
survfit2(Surv(time, nadm) ~ quantile, data = dapp1_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### CV event

```{r}
survfit2(Surv(time, mace) ~ quantile, data = dapp1_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Death

```{r}
survfit2(Surv(time, death_all) ~ quantile, data = dapp1_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Hazard ratios, CI, and p-values for Tables

```{r}
dapp1_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = dapp1_elisa)
tidy(dapp1_cox, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
dapp1_cox_nadm <- coxph(Surv(time, as.numeric(nadm)) ~ quantile, data = dapp1_elisa)
tidy(dapp1_cox_nadm, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
dapp1_cox_mace <- coxph(Surv(time, as.numeric(mace)) ~ quantile, data = dapp1_elisa)
tidy(dapp1_cox_mace, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
dapp1_cox_death <- coxph(Surv(time, as.numeric(death_all)) ~ quantile, data = dapp1_elisa)
tidy(dapp1_cox_death, conf.int = TRUE, exponentiate = TRUE)
```

### No. with event for Tables

```{r}
dapp1_elisa |> group_by(quantile) |> dplyr::count(group) |> filter(group == 1)
```

```{r}
dapp1_elisa |> group_by(quantile) |> dplyr::count(nadm) |> filter(nadm == 1)
```

```{r}
dapp1_elisa |> group_by(quantile) |> dplyr::count(mace) |> filter(mace == 1)
```

```{r}
dapp1_elisa |> group_by(quantile) |> dplyr::count(death_all) |> filter(death_all == 1)
```

## GAL9

```{r}
#| message: false
gal9_elisa <- read_tsv("../data/Results_GAL9_v2.tsv", locale = locale("es", decimal_mark = ",")) |> dplyr::select("Sample ID", "Mean Concentration [pg/ml]") |> rename_with(~ c("SampleID", "mean_conc")) |> 
  left_join((mdata |> dplyr::select(sid, olink_id, group, nadm, mace, death_all)), by = c("SampleID" = "olink_id")) |> 
  mutate(quantile = factor(ntile(mean_conc, 4))) |> 
  left_join(time_data, by = c("sid", "group")) |> 
  filter(time > 0) |> 
  mutate(group = as.factor(group))
```

```{r}
roc_res <- roc(gal9_elisa$group, gal9_elisa$mean_conc, direction = "<")  # use direction = "<" if higher values = higher risk
plot(roc_res, print.auc = TRUE)
```

```{r}
opt_cutoff <- coords(roc_res, "best", best.method = "youden", transpose = FALSE)
opt_cutoff
```

```{r}
survfit2(Surv(time, group) ~ quantile, data = gal9_elisa) |>
  # ggcuminc(linetype_aes = TRUE) +
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall time-to-event probability"
  ) +
  # scale_color_grey() +
  # scale_fill_grey() +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event") +
  ggtitle("Association between LGALS9 plasma concentrations and time to event")
```

```{r}
gal9_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = gal9_elisa)
summary(gal9_cox)
```

```{r}
quantile(gal9_elisa$mean_conc, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
```

```{r}
threshold_time <- 12*5  # in months
gal9_binary <- gal9_elisa |> 
  mutate(
    event_within_1y = ifelse(group == 1 & time <= threshold_time, 1, 0),
    test_positive = ifelse(quantile %in% c(1), 1, 0)
  )
```

```{r}
table(gal9_binary$test_positive, gal9_binary$event_within_1y)
```

```{r}
TP <- sum(gal9_binary$test_positive == 1 & gal9_binary$event_within_1y == 1, na.rm = TRUE)
FP <- sum(gal9_binary$test_positive == 1 & gal9_binary$event_within_1y == 0, na.rm = TRUE)
TN <- sum(gal9_binary$test_positive == 0 & gal9_binary$event_within_1y == 0, na.rm = TRUE)
FN <- sum(gal9_binary$test_positive == 0 & gal9_binary$event_within_1y == 1, na.rm = TRUE)
```

```{r}
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
PPV <- TP / (TP + FP)
NPV <- TN / (TN + FN)

results <- c(Sensitivity = sensitivity,
             Specificity = specificity,
             VPP = PPV,
             VPN = NPV)

round(results, 3)
```

5 years:

```{r}
survfit2(Surv(time_trunc, as.factor(group_trunc)) ~ quantile, data = gal9_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
summary(coxph(Surv(time_trunc, group_trunc) ~ quantile, data = gal9_elisa))
```

### Neoplasia

```{r}
survfit2(Surv(time, nadm) ~ quantile, data = gal9_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### CV event

```{r}
survfit2(Surv(time, mace) ~ quantile, data = gal9_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Death

```{r}
survfit2(Surv(time, death_all) ~ quantile, data = gal9_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Hazard ratios, CI, and p-values for Tables

```{r}
gal9_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = gal9_elisa)
tidy(gal9_cox, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
gal9_cox_nadm <- coxph(Surv(time, as.numeric(nadm)) ~ quantile, data = gal9_elisa)
tidy(gal9_cox_nadm, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
gal9_cox_mace <- coxph(Surv(time, as.numeric(mace)) ~ quantile, data = gal9_elisa)
tidy(gal9_cox_mace, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
gal9_cox_death <- coxph(Surv(time, as.numeric(death_all)) ~ quantile, data = gal9_elisa)
tidy(gal9_cox_death, conf.int = TRUE, exponentiate = TRUE)
```

### No. with event for Tables

```{r}
gal9_elisa |> group_by(quantile) |> dplyr::count(group) |> filter(group == 1)
```

```{r}
gal9_elisa |> group_by(quantile) |> dplyr::count(nadm) |> filter(nadm == 1)
```

```{r}
gal9_elisa |> group_by(quantile) |> dplyr::count(mace) |> filter(mace == 1)
```

```{r}
gal9_elisa |> group_by(quantile) |> dplyr::count(death_all) |> filter(death_all == 1)
```


## SKAP2

```{r}
#| message: false
skap2_elisa <- read_tsv("../data/Results_SKAP2_v2.tsv", locale = locale("es", decimal_mark = ",")) |> dplyr::select("Sample ID", "Mean Concentration [ng/ml]") |> rename_with(~ c("SampleID", "mean_conc")) |> 
  left_join((mdata |> dplyr::select(sid, olink_id, group, nadm, mace, death_all)), by = c("SampleID" = "olink_id")) |> 
  mutate(quantile = factor(ntile(mean_conc, 4))) |> 
  left_join(time_data, by = c("sid", "group")) |> 
  filter(time > 0) |> 
  mutate(group = as.factor(group))
```

```{r}
roc_res <- roc(skap2_elisa$group, skap2_elisa$mean_conc, direction = "<")  # use direction = "<" if higher values = higher risk
plot(roc_res, print.auc = TRUE)
```

```{r}
opt_cutoff <- coords(roc_res, "best", best.method = "youden", transpose = FALSE)
opt_cutoff
```

```{r}
survfit2(Surv(time, group) ~ quantile, data = skap2_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
skap2_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = skap2_elisa)
summary(skap2_cox)
```

```{r}
quantile(skap2_elisa$mean_conc, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
```

5 years:

```{r}
survfit2(Surv(time_trunc, as.factor(group_trunc)) ~ quantile, data = skap2_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

```{r}
summary(coxph(Surv(time_trunc, group_trunc) ~ quantile, data = skap2_elisa))
```

### Neoplasia

```{r}
survfit2(Surv(time, nadm) ~ quantile, data = skap2_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### CV event

```{r}
survfit2(Surv(time, mace) ~ quantile, data = skap2_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Death

```{r}
survfit2(Surv(time, death_all) ~ quantile, data = skap2_elisa) |>
  ggcuminc() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  # add_confidence_interval() +
  add_risktable(risktable_stats = "cum.event")
```

### Hazard ratios, CI, and p-values for Tables

```{r}
skap2_cox <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = skap2_elisa)
tidy(skap2_cox, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
skap2_cox_nadm <- coxph(Surv(time, as.numeric(nadm)) ~ quantile, data = skap2_elisa)
tidy(skap2_cox_nadm, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
skap2_cox_mace <- coxph(Surv(time, as.numeric(mace)) ~ quantile, data = skap2_elisa)
tidy(skap2_cox_mace, conf.int = TRUE, exponentiate = TRUE)
```

```{r}
skap2_cox_death <- coxph(Surv(time, as.numeric(death_all)) ~ quantile, data = skap2_elisa)
tidy(skap2_cox_death, conf.int = TRUE, exponentiate = TRUE)
```

### No. with event for Tables

```{r}
skap2_elisa |> group_by(quantile) |> dplyr::count(group) |> filter(group == 1)
```

```{r}
skap2_elisa |> group_by(quantile) |> dplyr::count(nadm) |> filter(nadm == 1)
```

```{r}
skap2_elisa |> group_by(quantile) |> dplyr::count(mace) |> filter(mace == 1)
```

```{r}
skap2_elisa |> group_by(quantile) |> dplyr::count(death_all) |> filter(death_all == 1)
```

## Final figure (4 panels)

```{r}
# Helper: minimally spread y positions to avoid label overlaps
spread_y <- function(y, min_sep = 0.03, y_min = 0, y_max = 1) {
  if (length(y) <= 1) return(y)

  ord <- order(y)
  ys  <- y[ord]

  # forward pass: enforce minimum separation
  for (i in 2:length(ys)) {
    if (ys[i] - ys[i - 1] < min_sep) ys[i] <- ys[i - 1] + min_sep
  }

  # if overflow, shift down and enforce backwards
  overflow <- max(ys) - y_max
  if (overflow > 0) {
    ys <- ys - overflow
    for (i in (length(ys) - 1):1) {
      if (ys[i + 1] - ys[i] < min_sep) ys[i] <- ys[i + 1] - min_sep
    }
  }

  ys <- pmin(pmax(ys, y_min), y_max)

  out <- y
  out[ord] <- ys
  out
}

# Plot function: one panel (curve + direct labels + risk table)
plot_cuminc_pub <- function(fit,
                            title,
                            xlab = "Months",
                            ylab = "Time-to-event prob.",
                            risktable_stats = "cum.event",
                            x_nudge = 0.6,         # horizontal offset for labels (in x-units, e.g. months)
                            min_sep = 0.03,        # vertical minimum separation between labels (in y-units)
                            right_expand = 0.2,   # extra space on right for labels
                            base_size = 10,
                            p_text = NULL) {

  # Base curve plot
  p <- ggcuminc(fit) +
    labs(x = xlab, y = ylab, color = "Quartile") +
    ggtitle(title) +
    theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0),
      axis.title = element_text(face = "bold"),
      legend.position = "none"   # remove legends (labels will identify curves)
    )

  # Get last point per curve from the built ggplot data
  lab_df <-
    p$data |>
    group_by(strata) |>
    slice_max(time, n = 1, with_ties = FALSE) |>
    ungroup() |>
    mutate(
      strata_chr = as.character(strata),
      qnum  = str_extract(strata_chr, "\\d+"),
      label = paste0("Q", qnum),
      x_label = time + x_nudge
    )

  # Spread y only if needed (minimal adjustment)
  lab_df <- lab_df |>
    mutate(y_label = spread_y(estimate, min_sep = min_sep, y_min = 0, y_max = 1))

  # Add direct labels (no leader lines)
  p <- p +
    geom_text(
      data = lab_df,
      aes(x = x_label, y = y_label, label = label, color = strata),
      inherit.aes = FALSE,
      fontface = "bold",
      hjust = 0,
      size = 3,
      show.legend = FALSE
    ) +
    coord_cartesian(clip = "off") +
    scale_x_continuous(expand = expansion(mult = c(0.02, right_expand)))
  
  if (!is.null(p_text)) {
  p <- p + annotate(
    "text",
    x = Inf, y = 0.02, 
    label = p_text,
    hjust = 1.1, vjust = 0,
    size = 3.2,
    fontface = "plain"
  )}

  # Add risk table under each panel
  p <- p + add_risktable(risktable_stats = risktable_stats)

  ggsurvfit_build(p)
}

# Fits
fit_gal9 <- survfit2(Surv(time, group) ~ quantile, data = gal9_elisa)
fit_ccl13 <- survfit2(Surv(time, group) ~ quantile, data = ccl13_elisa)
fit_dapp1 <- survfit2(Surv(time, group) ~ quantile, data = dapp1_elisa)
fit_skap2 <- survfit2(Surv(time, group) ~ quantile, data = skap2_elisa)

# Build each panel
p1 <- plot_cuminc_pub(fit_gal9, "LGALS9", p_text = "Global P value = 0.034")
p2 <- plot_cuminc_pub(fit_ccl13, "CCL13", p_text = "Global P value = 0.384")
p3 <- plot_cuminc_pub(fit_dapp1, "DAPP1", p_text = "Global P value = 0.359")
p4 <- plot_cuminc_pub(fit_skap2, "SKAP2", p_text = "Global P value = 0.906")

# Wrap each panel so patchwork tags only once (A–D), not the risk table separately
w1 <- wrap_elements(full = p1)
w2 <- wrap_elements(full = p2)
w3 <- wrap_elements(full = p3)
w4 <- wrap_elements(full = p4)

# Final figure
final_fig <-
  (w1 | w2) / (w3 | w4) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(face = "bold", size = 12),
    plot.tag.position = c(0.01, 0.99)
  )

# svg("../data/fig3AtoD.svg", width = 8, height = 8)
final_fig
# dev.off()
```

### NADM

```{r}
# Fits
fit_gal9 <- survfit2(Surv(time, nadm) ~ quantile, data = gal9_elisa)
fit_ccl13 <- survfit2(Surv(time, nadm) ~ quantile, data = ccl13_elisa)
fit_dapp1 <- survfit2(Surv(time, nadm) ~ quantile, data = dapp1_elisa)
fit_skap2 <- survfit2(Surv(time, nadm) ~ quantile, data = skap2_elisa)

# Build each panel
p1 <- plot_cuminc_pub(fit_gal9, "LGALS9", p_text = "Global P value = 0.578")
p2 <- plot_cuminc_pub(fit_ccl13, "CCL13", p_text = "Global P value = 0.643")
p3 <- plot_cuminc_pub(fit_dapp1, "DAPP1", p_text = "Global P value = 0.794")
p4 <- plot_cuminc_pub(fit_skap2, "SKAP2", p_text = "Global P value = 0.751")

# Wrap each panel so patchwork tags only once (A–D), not the risk table separately
w1 <- wrap_elements(full = p1)
w2 <- wrap_elements(full = p2)
w3 <- wrap_elements(full = p3)
w4 <- wrap_elements(full = p4)

# Final figure
final_fig <-
  (w1 | w2) / (w3 | w4) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(face = "bold", size = 12),
    plot.tag.position = c(0.01, 0.99)
  )

# svg("../data/figS2.svg", width = 8, height = 8)
final_fig
# dev.off()
```

### MACE

```{r}
# Fits
fit_gal9 <- survfit2(Surv(time, mace) ~ quantile, data = gal9_elisa)
fit_gal9_cont <- survfit2(Surv(time, mace) ~ as.numeric(quantile),
                          data = gal9_elisa)
fit_ccl13 <- survfit2(Surv(time, mace) ~ quantile, data = ccl13_elisa)
fit_dapp1 <- survfit2(Surv(time, mace) ~ quantile, data = dapp1_elisa)
fit_skap2 <- survfit2(Surv(time, mace) ~ quantile, data = skap2_elisa)

# Build each panel
p1 <- plot_cuminc_pub(fit_gal9, "LGALS9", p_text = "Global P value = 0.005")
p2 <- plot_cuminc_pub(fit_ccl13, "CCL13", p_text = "Global P value = 0.385")
p3 <- plot_cuminc_pub(fit_dapp1, "DAPP1", p_text = "Global P value = 0.052")
p4 <- plot_cuminc_pub(fit_skap2, "SKAP2", p_text = "Global P value = 0.790")

# Wrap each panel so patchwork tags only once (A–D), not the risk table separately
w1 <- wrap_elements(full = p1)
w2 <- wrap_elements(full = p2)
w3 <- wrap_elements(full = p3)
w4 <- wrap_elements(full = p4)

# Final figure
final_fig <-
  (w1 | w2) / (w3 | w4) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(face = "bold", size = 12),
    plot.tag.position = c(0.01, 0.99)
  )

# svg("../data/figS3.svg", width = 8, height = 8)
final_fig
# dev.off()
```


### Death

```{r}
# Fits
fit_gal9 <- survfit2(Surv(time, death_all) ~ quantile, data = gal9_elisa)
fit_ccl13 <- survfit2(Surv(time, death_all) ~ quantile, data = ccl13_elisa)
fit_dapp1 <- survfit2(Surv(time, death_all) ~ quantile, data = dapp1_elisa)
fit_skap2 <- survfit2(Surv(time, death_all) ~ quantile, data = skap2_elisa)

# Build each panel
p1 <- plot_cuminc_pub(fit_gal9, "LGALS9", p_text = "Global P value = 0.025")
p2 <- plot_cuminc_pub(fit_ccl13, "CCL13", p_text = "Global P value = 0.436")
p3 <- plot_cuminc_pub(fit_dapp1, "DAPP1", p_text = "Global P value = 0.954")
p4 <- plot_cuminc_pub(fit_skap2, "SKAP2", p_text = "Global P value = 0.518")

# Wrap each panel so patchwork tags only once (A–D), not the risk table separately
w1 <- wrap_elements(full = p1)
w2 <- wrap_elements(full = p2)
w3 <- wrap_elements(full = p3)
w4 <- wrap_elements(full = p4)

# Final figure
final_fig <-
  (w1 | w2) / (w3 | w4) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(face = "bold", size = 12),
    plot.tag.position = c(0.01, 0.99)
  )

# svg("../data/figS4.svg", width = 8, height = 8)
final_fig
# dev.off()
```

## Assessing linearity of the association between HR and protein concentration

### GAL9

Transform to log2 scale:

```{r}
gal9_elisa <- gal9_elisa %>%
  mutate(log2_gal9 = log2(mean_conc))

gal9_elisa$event_group <- ifelse(gal9_elisa$group == "1", 1, 0)
```

```{r}
cox_pspline <- coxph(
  Surv(time, event_group) ~ pspline(log2_gal9, df = 3),
  data = gal9_elisa
)
```

```{r}
newdat <- data.frame(log2_gal9 = seq(min(gal9_elisa$log2_gal9),
                                     max(gal9_elisa$log2_gal9),
                                      length = 200))
```

```{r}
pred <- predict(cox_pspline, newdata = newdat, type = "risk", se.fit = TRUE)
```

```{r}
plot(newdat$log2_gal9, pred$fit, type = "l",
     xlab = "log2(GAL9)", ylab = "Relative Hazard")
```

```{r}
# Prepare data frame for plotting
plotdat <- newdat %>%
  mutate(
    HR = exp(pred$fit),
    HR_low = exp(pred$fit - 1.96 * pred$se.fit),
    HR_high = exp(pred$fit + 1.96 * pred$se.fit),
    GAL9 = 2^log2_gal9  # back-transform to original scale
  )

q_vals <- quantile(gal9_elisa$log2_gal9, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)

# ggplot
# ggplot(plotdat, aes(x = GAL9, y = HR)) +
p1 <- ggplot(plotdat, aes(x = log2_gal9, y = HR)) +  
  geom_ribbon(aes(ymin = HR_low, ymax = HR_high), alpha = 0.25) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = q_vals, linetype = "dashed", color = "black") +
  geom_rug(data = gal9_elisa, aes(x = log2_gal9), sides = "b", alpha = 0.4,
           inherit.aes = FALSE) +
  scale_x_continuous("log2 GAL9") +
  scale_y_continuous("Relative Hazard") +
  theme_minimal(base_size = 14)
```

### DAPP1

Transform to log2 scale:

```{r}
dapp1_elisa <- dapp1_elisa %>%
  mutate(log2_dapp1 = log2(mean_conc))

dapp1_elisa$event_group <- ifelse(dapp1_elisa$group == "1", 1, 0)
```

```{r}
cox_pspline <- coxph(
  Surv(time, event_group) ~ pspline(log2_dapp1, df = 3),
  data = dapp1_elisa
)
```

```{r}
newdat <- data.frame(log2_dapp1 = seq(min(dapp1_elisa$log2_dapp1),
                                     max(dapp1_elisa$log2_dapp1),
                                      length = 200))
```

```{r}
pred <- predict(cox_pspline, newdata = newdat, type = "risk", se.fit = TRUE)
```

```{r}
plot(newdat$log2_dapp1, pred$fit, type = "l",
     xlab = "log2(DAPP1)", ylab = "Relative Hazard")
```

```{r}
# Prepare data frame for plotting
plotdat <- newdat %>%
  mutate(
    HR = exp(pred$fit),
    HR_low = exp(pred$fit - 1.96 * pred$se.fit),
    HR_high = exp(pred$fit + 1.96 * pred$se.fit),
    DAPP1 = 2^log2_dapp1  # back-transform to original scale
  )

q_vals <- quantile(dapp1_elisa$log2_dapp1, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)

# ggplot
# ggplot(plotdat, aes(x = GAL9, y = HR)) +
p2 <- ggplot(plotdat, aes(x = log2_dapp1, y = HR)) +  
  geom_ribbon(aes(ymin = HR_low, ymax = HR_high), alpha = 0.25) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = q_vals, linetype = "dashed", color = "black") +
  geom_rug(data = dapp1_elisa, aes(x = log2_dapp1), sides = "b", alpha = 0.4,
           inherit.aes = FALSE) +
  scale_x_continuous("log2 DAPP1") +
  scale_y_continuous("Relative Hazard") +
  theme_minimal(base_size = 14)
```

### CCL13

Transform to log2 scale:

```{r}
ccl13_elisa <- ccl13_elisa %>%
  mutate(log2_ccl13 = log2(mean_conc)) |> 
  filter(!is.na(mean_conc))

ccl13_elisa$event_group <- ifelse(ccl13_elisa$group == "1", 1, 0)
```

```{r}
cox_pspline <- coxph(
  Surv(time, event_group) ~ pspline(log2_ccl13, df = 3),
  data = ccl13_elisa
)
```

```{r}
newdat <- data.frame(log2_ccl13 = seq(min(ccl13_elisa$log2_ccl13),
                                     max(ccl13_elisa$log2_ccl13),
                                      length = 200))
```

```{r}
pred <- predict(cox_pspline, newdata = newdat, type = "risk", se.fit = TRUE)
```

```{r}
plot(newdat$log2_ccl13, pred$fit, type = "l",
     xlab = "log2(CCL13)", ylab = "Relative Hazard")
```

```{r}
# Prepare data frame for plotting
plotdat <- newdat %>%
  mutate(
    HR = exp(pred$fit),
    HR_low = exp(pred$fit - 1.96 * pred$se.fit),
    HR_high = exp(pred$fit + 1.96 * pred$se.fit),
    CCL13 = 2^log2_ccl13  # back-transform to original scale
  )

q_vals <- quantile(ccl13_elisa$log2_ccl13, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)

# ggplot
# ggplot(plotdat, aes(x = GAL9, y = HR)) +
p3 <- ggplot(plotdat, aes(x = log2_ccl13, y = HR)) +  
  geom_ribbon(aes(ymin = HR_low, ymax = HR_high), alpha = 0.25) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = q_vals, linetype = "dashed", color = "black") +
  geom_rug(data = ccl13_elisa, aes(x = log2_ccl13), sides = "b", alpha = 0.4,
           inherit.aes = FALSE) +
  scale_x_continuous("log2 CCL13") +
  scale_y_continuous("Relative Hazard") +
  theme_minimal(base_size = 14)
```

### SKAP2

Transform to log2 scale:

```{r}
skap2_elisa <- skap2_elisa %>%
  mutate(log2_skap2 = log2(mean_conc)) |> 
  filter(!is.na(mean_conc))

skap2_elisa$event_group <- ifelse(skap2_elisa$group == "1", 1, 0)
```

```{r}
cox_pspline <- coxph(
  Surv(time, event_group) ~ pspline(log2_skap2, df = 3),
  data = skap2_elisa
)
```

```{r}
newdat <- data.frame(log2_skap2 = seq(min(skap2_elisa$log2_skap2),
                                     max(skap2_elisa$log2_skap2),
                                      length = 200))
```

```{r}
pred <- predict(cox_pspline, newdata = newdat, type = "risk", se.fit = TRUE)
```

```{r}
plot(newdat$log2_skap2, pred$fit, type = "l",
     xlab = "log2(SKAP2)", ylab = "Relative Hazard")
```

```{r}
# Prepare data frame for plotting
plotdat <- newdat %>%
  mutate(
    HR = exp(pred$fit),
    HR_low = exp(pred$fit - 1.96 * pred$se.fit),
    HR_high = exp(pred$fit + 1.96 * pred$se.fit),
    SKAP2 = 2^log2_skap2  # back-transform to original scale
  )

q_vals <- quantile(skap2_elisa$log2_skap2, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)

# ggplot
# ggplot(plotdat, aes(x = GAL9, y = HR)) +
p4 <- ggplot(plotdat, aes(x = log2_skap2, y = HR)) +  
  geom_ribbon(aes(ymin = HR_low, ymax = HR_high), alpha = 0.25) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = q_vals, linetype = "dashed", color = "black") +
  geom_rug(data = skap2_elisa, aes(x = log2_skap2), sides = "b", alpha = 0.4,
           inherit.aes = FALSE) +
  scale_x_continuous("log2 SKAP2") +
  scale_y_continuous("Relative Hazard") +
  theme_minimal(base_size = 14)
```

### 4-panel figure

```{r}
# svg("../data/FigS2.svg", width = 7, height = 7)
ggarrange(p1, p2, p3, p4, labels = "AUTO")
# dev.off()
```

## Assessment of proportional hazards assumption using Schoenfeld residuals

### Plotting function

```{r}
plot_schoenfeld_gg <- function(fit,
                               transform = "km",
                               xlab = "Time (KM scale)",
                               ylab = "Scaled Schoenfeld residuals",
                               title = NULL) {
  
  ph <- cox.zph(fit, transform = transform)

  df <- data.frame(
    time  = ph$x,
    resid = ph$y[, 1]   # only one covariate → first column
  )

  pval <- ph$table[1, "p"]

  ggplot(df, aes(time, resid)) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
    geom_point(alpha = 0.4, size = 1) +
    geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 0.8) +
    labs(
      x = xlab,
      y = ylab,
      title = title,
      subtitle = paste0(
        "Proportional hazards test: p = ",
        formatC(pval, digits = 3, format = "f")
      )
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = 10)
    )
}
```


### GAL9

```{r}
cox_gal9 <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = gal9_elisa)
ph_gal9 <- cox.zph(cox_gal9)
ph_gal9
```

```{r}
s1 <- plot_schoenfeld_gg(cox_gal9,
                   transform = "km",
                   xlab = "Time (K-M transform)",
                   title = "LGALS9")
```

### DAPP1

```{r}
cox_dapp1 <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = dapp1_elisa)
ph_dapp1 <- cox.zph(cox_dapp1)
ph_dapp1
```

```{r}
s2 <- plot_schoenfeld_gg(cox_dapp1,
                   transform = "km",
                   xlab = "Time (K-M transform)",
                   title = "DAPP1")
```

### CCL13

```{r}
cox_ccl13 <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = ccl13_elisa)
ph_ccl13 <- cox.zph(cox_ccl13)
ph_ccl13
```

```{r}
s3 <- plot_schoenfeld_gg(cox_ccl13,
                   transform = "km",
                   xlab = "Time (K-M transform)",
                   title = "CCL13")
```

### SKAP2

```{r}
cox_skap2 <- coxph(Surv(time, as.numeric(group)) ~ quantile, data = skap2_elisa)
ph_skap2 <- cox.zph(cox_skap2)
ph_skap2
```

```{r}
s4 <- plot_schoenfeld_gg(cox_skap2,
                   transform = "km",
                   xlab = "Time (K-M transform)",
                   title = "SKAP2")
```

### 4-panel figure

```{r}
# svg("../data/FigS6.svg", width = 7, height = 7)
ggarrange(s1, s2, s3, s4, labels = "AUTO")
# dev.off()
```

