---
title: "Correlation analysis"
format:
  html:
    df-print: paged
    embed-resources: true
---

```{r}
#| message: false
library(haven)
library(OlinkAnalyze)
library(readxl)
library(tidyverse)
```

Import proteomics data:

```{r}
#| message: false
pdata <- read_tsv("../data/PRETTY_NPX_138samp_326prot_QCed.tsv")
```

Import metadata:

```{r}
#| message: false
mdata <- pdata |>
  dplyr::select(sid:death_all) |>
  distinct() |>
  arrange(sid)
mdata_rnaseq <- read_tsv("../data/PRETTY_metadata.tsv") |> 
  dplyr::select(sid, rnaseq_id, group)
```

Read differentially expressed proteins:

```{r}
#| message: false
dep_group <- read_csv("../data/dep_group.csv") |> 
  filter(Threshold == "Significant") |> pull(Assay)
dep_nadm <- read_csv("../data/dep_nadm.csv") |> 
  filter(Threshold == "Significant") |> pull(Assay)
dep_mace <- read_csv("../data/dep_mace.csv") |> 
  filter(Threshold == "Significant") |> pull(Assay)
dep_death_all <- read_csv("../data/dep_death_all.csv") |> 
  filter(Threshold == "Significant") |> pull(Assay)
dep_all <- unique(c(dep_group, dep_nadm, dep_mace, dep_death_all))
```

Read differentially expressed genes:

```{r}
#| message: false
deg_group <- read_csv("../data/deg_sig_group.csv") |> dplyr::select(gene, external_gene_name) |> filter(!is.na(external_gene_name))
deg_nadm <- read_csv("../data/deg_sig_nadm.csv") |> dplyr::select(gene, external_gene_name) |> filter(!is.na(external_gene_name))
deg_mace <- read_csv("../data/deg_sig_mace.csv") |> dplyr::select(gene, external_gene_name) |> filter(!is.na(external_gene_name))
deg_death_all <- read_csv("../data/deg_sig_death_all.csv") |> dplyr::select(gene, external_gene_name) |> filter(!is.na(external_gene_name))
deg_all <- rbind(deg_group, deg_nadm, deg_mace, deg_death_all) |> distinct()
```

Subset proteomics data:

```{r}
pdata_cors <- pdata |> 
  filter(Assay %in% dep_all) |> 
  dplyr::select(sid, Assay, NPX) |> 
  pivot_wider(names_from = Assay, values_from = NPX) |> 
  arrange(sid)
```

Subset transcriptomics data:

```{r}
#| message: false
rdata_cors <- read_csv("../data/vst_normalized_counts.csv") |>
  filter(gene %in% deg_all$gene) |> 
  dplyr::select(-1) |> 
  pivot_longer(cols = P_1:P_99, names_to = "rnaseq_id", values_to = "vst")
```

Change the ids of the transcriptomics data to match the proteomics data:

```{r}
id_pbmc <- read_xlsx("../data/id_group_PBMC.xlsx") |> 
  mutate(rnaseqid = gsub("-", "_", IDTUBO)) |> 
  dplyr::select(id_sample, rnaseqid) |> 
  left_join(mdata_rnaseq, by = c("rnaseqid" = "rnaseq_id")) |> 
  dplyr::select(sid, rnaseqid)
rdata_cors <- rdata_cors |> 
  left_join(id_pbmc, by = c("rnaseq_id" = "rnaseqid")) |> 
  dplyr::select(sid, gene, vst) |> 
  pivot_wider(names_from = gene, values_from = vst) |> 
  filter(sid %in% pdata_cors$sid) |> 
  arrange(sid)
```

Check that all participants in the proteomics data have got transcriptomics data:

```{r}
pdata_cors <- pdata_cors |> filter(sid %in% rdata_cors$sid)
```

Standardize variables:

```{r}
pdata_cors <- pdata_cors |> mutate(across(-1, ~ scale(.)[, 1]))
rdata_cors <- rdata_cors |> mutate(across(-1, ~ scale(.)[, 1]))
```

Correlations:

```{r}
rdata_cors <- rdata_cors[, -1]
pdata_cors <- pdata_cors[, -1]

pairs <- expand.grid(
  RNA = names(rdata_cors),
  Protein = names(pdata_cors),
  stringsAsFactors = FALSE
)

results <- mapply(function(rna_col, prot_col) {
  x <- rdata_cors[[rna_col]]
  y <- pdata_cors[[prot_col]]
  
  test <- cor.test(x, y, use = "complete.obs")
  
  data.frame(
    RNA = rna_col,
    Protein = prot_col,
    Correlation = test$estimate,
    P.value = test$p.value
  )
}, pairs$RNA, pairs$Protein, SIMPLIFY = FALSE)

cor_results_df <- do.call(rbind, results)
cor_results_df$adj.p.value <- p.adjust(cor_results_df$P.value, method = "fdr")
cor_results_df <- cor_results_df |> 
  mutate(signif = case_when(
        adj.p.value < 0.001 ~ "**",
        adj.p.value < 0.01  ~ "*",
        adj.p.value < 0.05  ~ ".",
        TRUE            ~ ""
  ))
cor_results_df |> filter(adj.p.value < 0.05) |> arrange(Correlation)
```

Add names to RNA ENSEMBL IDs:

```{r}
cor_results_df <- cor_results_df |>
  left_join(deg_all, by = c("RNA" = "gene"))
```

```{r}
cor_results_df |> 
  filter(adj.p.value < 0.05) |> 
  arrange(Correlation) |> 
  dplyr::count(Protein) |> 
  arrange(n) |> 
  filter(Protein %in% c("CCL13", "CLIP2", "DAPP1", "FCAR", "GAL9", "IL6", "SKAP2", "SPINK4", "TNF")) |> arrange(desc(n))
```

```{r}
cor_results_df |> 
    filter(adj.p.value < 0.05) |> 
    arrange(Correlation) |> 
    dplyr::count(RNA) |> 
    arrange(desc(n)) |> 
  left_join(deg_all, by = c("RNA" = "gene"))
```

Visualization:

```{r}
ggplot(cor_results_df, aes(x = Protein, y = external_gene_name, fill = Correlation)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                         midpoint = 0, limit = c(-1, 1), 
                         name = "Correlation") +
    geom_text(aes(label = signif), size = 5, color = "black") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Correlation Matrix Heatmap")
```

```{r}
mat_df <- cor_results_df |>
  dplyr::select(external_gene_name, Protein, Correlation) |> 
  pivot_wider(names_from = Protein, values_from = Correlation)
mat <- mat_df |> 
  dplyr::select(-1) |> 
  as.matrix()
rownames(mat) <- mat_df$external_gene_name

sig_mat_df <- cor_results_df |>
  mutate(significant = ifelse(adj.p.value < 0.05, 1, 0)) |> 
  dplyr::select(external_gene_name, Protein, significant) |> 
  pivot_wider(names_from = Protein, values_from = significant)
sig_mat <- sig_mat_df |> 
  dplyr::select(-1) |> 
  as.matrix()
rownames(sig_mat) <- mat_df$external_gene_name
```

```{r}
library(ComplexHeatmap)
library(circlize)
library(grid)

ht <- Heatmap(
  mat,
  name = "Correlation",
  column_title = "Differentially expressed proteins",
  row_title = "Differentially expressed genes",
  column_title_side = "bottom",
  row_title_side = "right",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  row_title_gp = gpar(fontsize = 12, fontface = "bold"),
  na_col = "grey90",
  
  # Cell function
    cell_fun = function(j, i, x, y, w, h, fill) {
    if (sig_mat[i, j]) {
      grid::grid.rect(
        x = x, y = y,
        width = w, height = h,
        gp = grid::gpar(col = "black", fill = NA, lwd = 1)
      )
    }
  },

  # Clustering (publication-friendly defaults)
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",

  # Dendrogram control
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_dend_width = unit(25, "mm"),
  column_dend_height = unit(25, "mm"),

  # Labels / text
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9),
  column_names_rot = 45,

  # Cell borders (subtle)
  rect_gp = gpar(col = "grey85", lwd = 0.6),

  # Legend styling
  heatmap_legend_param = list(
    title = "Spearman\ncorrelation",
    at = c(-1, -0.5, 0, 0.5, 1),
    labels = c("-1", "-0.5", "0", "0.5", "1"),
    legend_height = unit(45, "mm"),
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 9)
  )
)

# svg("../data/correlation_heatmap.svg", width = 14, height = 9)
draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  padding = unit(c(4, 4, 4, 4), "mm"),
)
# dev.off()

```

## Correlation analysis: only SNAEs

Subset proteomics data:

```{r}
pdata_cors <- pdata |> 
  filter(Assay %in% dep_group) |> 
  dplyr::select(sid, Assay, NPX) |> 
  pivot_wider(names_from = Assay, values_from = NPX) |> 
  arrange(sid)
```

Subset transcriptomics data:

```{r}
#| message: false
rdata_cors <- read_csv("../data/vst_normalized_counts.csv") |>
  filter(gene %in% deg_group$gene) |> 
  dplyr::select(-1) |> 
  pivot_longer(cols = P_1:P_99, names_to = "rnaseq_id", values_to = "vst")
```

Change the ids of the transcriptomics data to match the proteomics data:

```{r}
id_pbmc <- read_xlsx("../data/id_group_PBMC.xlsx") |> 
  mutate(rnaseqid = gsub("-", "_", IDTUBO)) |> 
  dplyr::select(id_sample, rnaseqid) |> 
  left_join(mdata_rnaseq, by = c("rnaseqid" = "rnaseq_id")) |> 
  dplyr::select(sid, rnaseqid)
rdata_cors <- rdata_cors |> 
  left_join(id_pbmc, by = c("rnaseq_id" = "rnaseqid")) |> 
  dplyr::select(sid, gene, vst) |> 
  pivot_wider(names_from = gene, values_from = vst) |> 
  filter(sid %in% pdata_cors$sid) |> 
  arrange(sid)
```

Check that all participants in the proteomics data have got transcriptomics data:

```{r}
pdata_cors <- pdata_cors |> filter(sid %in% rdata_cors$sid)
```

Standardize variables:

```{r}
pdata_cors <- pdata_cors |> mutate(across(-1, ~ scale(.)[, 1]))
rdata_cors <- rdata_cors |> mutate(across(-1, ~ scale(.)[, 1]))
```

Correlations:

```{r}
rdata_cors <- rdata_cors[, -1]
pdata_cors <- pdata_cors[, -1]

pairs <- expand.grid(
  RNA = names(rdata_cors),
  Protein = names(pdata_cors),
  stringsAsFactors = FALSE
)

results <- mapply(function(rna_col, prot_col) {
  x <- rdata_cors[[rna_col]]
  y <- pdata_cors[[prot_col]]
  
  test <- cor.test(x, y, use = "complete.obs")
  
  data.frame(
    RNA = rna_col,
    Protein = prot_col,
    Correlation = test$estimate,
    P.value = test$p.value
  )
}, pairs$RNA, pairs$Protein, SIMPLIFY = FALSE)

cor_results_df <- do.call(rbind, results)
cor_results_df$adj.p.value <- p.adjust(cor_results_df$P.value, method = "fdr")
cor_results_df <- cor_results_df |> 
  mutate(signif = case_when(
        adj.p.value < 0.001 ~ "**",
        adj.p.value < 0.01  ~ "*",
        adj.p.value < 0.05  ~ ".",
        TRUE            ~ ""
  ))
cor_results_df |> filter(adj.p.value < 0.05) |> arrange(Correlation)
```

Add names to RNA ENSEMBL IDs:

```{r}
cor_results_df <- cor_results_df |>
  left_join(deg_all, by = c("RNA" = "gene"))
```

```{r}
cor_results_df |> 
  filter(adj.p.value < 0.05) |> 
  arrange(Correlation) |> 
  dplyr::count(Protein) |> 
  arrange(n) |> 
  filter(Protein %in% c("CCL13", "CLIP2", "DAPP1", "FCAR", "GAL9", "IL6", "SKAP2", "SPINK4", "TNF")) |> arrange(desc(n))
```

```{r}
cor_results_df |> 
    filter(adj.p.value < 0.05) |> 
    arrange(Correlation) |> 
    dplyr::count(RNA) |> 
    arrange(desc(n)) |> 
  left_join(deg_all, by = c("RNA" = "gene"))
```

Visualization:

```{r}
ggplot(cor_results_df, aes(x = Protein, y = external_gene_name, fill = Correlation)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                         midpoint = 0, limit = c(-1, 1), 
                         name = "Correlation") +
    geom_text(aes(label = signif), size = 5, color = "black") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Correlation Matrix Heatmap")
```

```{r}
mat_df <- cor_results_df |>
  dplyr::select(external_gene_name, Protein, Correlation) |> 
  pivot_wider(names_from = Protein, values_from = Correlation)
mat <- mat_df |> 
  dplyr::select(-1) |> 
  as.matrix()
rownames(mat) <- mat_df$external_gene_name

sig_mat_df <- cor_results_df |>
  mutate(significant = ifelse(adj.p.value < 0.05, 1, 0)) |> 
  dplyr::select(external_gene_name, Protein, significant) |> 
  pivot_wider(names_from = Protein, values_from = significant)
sig_mat <- sig_mat_df |> 
  dplyr::select(-1) |> 
  as.matrix()
rownames(sig_mat) <- mat_df$external_gene_name
```

```{r}
ht <- Heatmap(
  mat,
  name = "Correlation",
  column_title = "Differentially expressed proteins",
  row_title = "Differentially expressed genes",
  column_title_side = "bottom",
  row_title_side = "right",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  row_title_gp = gpar(fontsize = 12, fontface = "bold"),
  na_col = "grey90",
  
  # Cell function
    cell_fun = function(j, i, x, y, w, h, fill) {
    if (sig_mat[i, j]) {
      grid::grid.rect(
        x = x, y = y,
        width = w, height = h,
        gp = grid::gpar(col = "black", fill = NA, lwd = 1)
      )
    }
  },

  # Clustering (publication-friendly defaults)
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",

  # Dendrogram control
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_dend_width = unit(25, "mm"),
  column_dend_height = unit(25, "mm"),

  # Labels / text
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9),
  column_names_rot = 45,

  # Cell borders (subtle)
  rect_gp = gpar(col = "grey85", lwd = 0.6),

  # Legend styling
  heatmap_legend_param = list(
    title = "Spearman\ncorrelation",
    at = c(-1, -0.5, 0, 0.5, 1),
    labels = c("-1", "-0.5", "0", "0.5", "1"),
    legend_height = unit(45, "mm"),
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 9)
  )
)

# svg("../data/correlation_heatmap_group.svg", width = 10, height = 5)
draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  padding = unit(c(4, 4, 4, 4), "mm"),
)
# dev.off()
```
